Jane, John … Leslie? A Historical Method for Algorithmic Gender Prediction

### 2015-06-23 article.Rmd
submission files
First crack at revisions
Rewrote abstract, introduction, started to change AHR section

---
title: "Jane, John … Leslie? A Historical Method for Algorithmic Gender Prediction"
author: "Cameron Blevins and Lincoln Mullen"
output:
  html_document: default
---

@@

This paper addresses two trends in the humanities: the use of gender as an analytical framework and the proliferation of large, machine-readable datasets. It outlines a new method for algorithmically inferring the gender of a name using historical data. In contrast to existing methods of predicting gender, this method takes into account how naming practices change over time and space. It provides a higher level of accuracy and a precise statement of the likelihood that a name was a given gender, especially for datasets where names can be associated with dates. The article describes the methodology as implemented in the [Gender](https://github.com/ropensci/gender) package for the [R programming language](http://www.r-project.org/). It goes on to apply the method to a case study in which we examine gender and gatekeeping in the history profession by inferring the gender of authors of history dissertations and of authors and reviewers in the *American Historical Review*. The Gender package illustrates how humanities scholars can make concrete methodologial contributions to computer science and related fields.

@@

As humanities researchers grapple with larger and larger datasets they will increasingly be able to not just extract existing information from that data but to also infer additional information from it [Goldstone and Underwood 2014].^[The availability of large-scale datasets dealing with personal information raises important ethical issues for digital humanists, particularly in light of the revelations about domestic spying at the NSA brought forth by Edward Snowden in 2013. [Weingart 2014] Kieran Healy has used network analysis to demonstrate how researchers can infer information from datasets while also touching on the troubling ethical issue of surveillance [Healy 2013].] Many of these datasets deal with lists of people: company payrolls, military rosters, passenger bookings, records of correspondence, or lists of published works. Some of these records might contain associated information such as a person's address, age, or rank, while others contain little more than a list of personal names. A common question in the history of American religion, for example, is whether more women than men participated in a church or religious group. A simple membership roll might contain names, but it is unlikely to identify the gender of the members. One might also ask a question about changing patterns in male and female authorship of books. Library catalogs contain enormous amounts of data about publication, but they do not record the sex of authors.

Given large but sparse datasets, researchers have to use their ingenuity to derive additional kinds of information. What can researchers infer from just a name? On a basic level, first names often imply a person's gender. "Jane Fay" is almost certainly a woman, while her brother "John Fay" is almost certainly a man. Given a dataset matching names to genders, a straightforward computer program can easily infer their respective genders.

The problem with inferring gender from names is that the link between gender and naming practices, like language itself, is not static. What about Jane and John Fay's sibling, Leslie? If Leslie was born in the past sixty years, chances are good that Leslie would be their sister. But if the three siblings were born in the early twentieth century, chances are good that Leslie would be their brother. That is because the conventional gender for the name Leslie switched over the course of one hundred years. In 1900 some 92% of the babies born in the United States who were named Leslie were male, while in 2000 about 96% of the Leslies born in that year were female.

@@

Other names used in the United States such as Madison, Morgan, Sydney, and Kendall, have also flipped their gender over the past century. Such changes usually follow set patterns. In general, names tend to shift from being predominantly male to being predominantly female; the shift rarely happens in the other direction. In general, names that transition from male to female also became increasingly popular over time. For example, the name Madison changed from male to female quite suddenly after 1985. Whereas before it had been an infrequently used male name, after 1985 it skyrocketed in popularity and by the early 2000s was one of the ten most frequently used names for baby girls. Broadly speaking, in the late twentieth century Americans girls became increasingly more likely than American boys to receive a "trendy" name, and once that name was associated with women it was unlikely to be given to male babies. These names present a moving target for researchers hoping to use them to infer additional information about the people under study. In addition, the prevalence of changes in the gender of names in the last half of the twentieth century implies that gendered lists of names drawn from contemporary data are anachronistic and so of little use for earlier periods.^[The few exceptions of names which flip from being female to male, such as "Jan", have a different causal mechanism. We speculate that Jan as an English name is female but that Jan as a Scandinavian name is male. Because the name has remained relatively uncommon, other causes besides changing gender preferences can explain the change.]

@@

For those working on contemporary subjects the "Leslie problem" is not an especially pressing one. An education policy analyst studying the demographics of different urban school districts, for instance, can turn to a variety of tools that use current databases of names, and these tools will likely identify the gender of Leslie Fay correctly. One online service, [Genderize.io](http://genderize.io/), determines the gender of first names "from user profiles across major social networks." [Genderize.io n.d.] Checking the service for the name Leslie returns a prediction that the name is female, along with an estimate of male/female proportions of the uses among members of unnamed social media sites.^[In this example the Genderize.io service has been accessed through an API wrapper provided in our Gender package described below.]

```{r genderize, echo=TRUE, results='markup'}
gender("Leslie", method = "genderize")
```

But researchers studying a longer timespan need to take into account the changing nature of naming practices. If that same education researcher wanted to compare the demographics of contemporary urban schools with those from the early twentieth century, existing tools like Genderize.io would misidentify Leslie Fay's gender in historical datasets. The "Leslie problem" is not just for researchers who think of themselves as historians. Anyone studying a period longer than a few years, or anyone studying a group whose demographics do not match the groups used by a contemporary tool such as Genderize.io will also encounter this problem. As of 2012, the average American lifespan was nearly seventy-nine years---more than enough time for naming practices to change quite dramatically between their births and deaths [World Bank 2014]. Predicting gender from first names therefore requires a fundamentally historical method.

Our solution to the "Leslie problem" is to create a software package that combines a predictive algorithm with several historical datasets suitable to various times and regions. The algorithm calculates the proportion of male or female uses of a name in a given birth year or range of years. It thus can provide not only a prediction of the gender of a name, but also a measure of that prediction's likelihood. For example, our program predicts that a person named Leslie born in 1950 is female, but only just barely: 52 percent of the babies named Leslie born in the United States in 1950 were girls and 48 percent were boys. Including these probability figures allows researchers to conduct more nuanced analyses and to determine for themselves what level of certainty is acceptable for predicting the gender of names.

We recognize that gender is a fluid concept, both today and in the past, and that this method glosses over the complexities of historical definitions of gender and of the relationship between biological sex and gender [Butler 1990, Butler 1993]. In some sense this is unavoidable since the large historical datasets which make this kind of analysis possible were created by state agencies such as the U.S. Census Bureau. For the category of gender (as for race) states have offered a limited number of boxes to check, and those boxes exert a tremendous power to define state subjects [Brubaker and Cooper 2000, 14-17, Pascoe 2009, Canaday 2011, Landrum 2014]. This method does not presuppose that the gender binary imposed by governments and their wider societies reflect how an individual identifies oneself and is neutral as to what the categories "male" or "female" actually mean in any given historical context. Researchers are free---indeed, obligated---to make that determination for themselves, and this package provides access to the underlying dataset so that researchers can interrogate every assumption that comes with it. This method does not argue against the complexities of historical understandings of gender, but as with any project it does require researchers who use the method to think through how the data was gathered, what it implies, and how it reflects social and cultural practices.

The remainder of this article is divided into two sections. First, we describe our method in more detail, compare it to existing methods, and explain how to use it. Second, we apply the method to a case study of gatekeeping in the historical profession in order to demonstrate its usefulness.

@@

In a 2005 report for the American Historical Association, Elizabeth Lunbeck acknowledged "a sea change in the [historical] profession with respect to gender" before going on to describe the painful limitations of this sea change for female historians: ongoing personal discrimination, lower salaries, and barriers to securing high-ranking positions. Lunbeck's report drew in part on a survey of 362 female historians that produced a rich source of responses detailing the deep and multi-faceted challenges facing women in the profession [Lunbeck 2005]. What follows is a quantitative supplement to Lunbeck's analysis that uses our program to analyze gender representation amongst historians across a much larger scale and a much longer time period. We have found that although women have achieved rough equality in terms of the number of dissertations written, there is still a significant gap between male and female authorship of books in history.

Our analysis focuses on one of the bedrocks of the historical profession: scholarly research. We begin with the history dissertation, often the defining scholarly output of a historian's early career. The completion of a dissertation marks a standardized moment of transition out of the "training" phase of the historical profession. To study dissertations about history, we used data supplied by ProQuest for roughly eighty thousand PhD dissertations completed in the United States.^[ProQuest's dataset does not permit us to limit dissertations just to dissertations produced in history departments. These dissertations include dissertations *about* history. For a justification of this approach, see a [series of blog posts](http://lincolnmullen.com/research/history-dissertations/) exploring the dataset. The American Historical Association's dataset of dissertations relies on reporting from departments and does not have the chronological range of the ProQuest data. Nevertheless, applying the same method to the AHA's dataset does not provide any reason for thinking that the proportion of authorship by gender over time is substantially different when limited to just history department dissertations. American Historical Association, *Directory of History Dissertations*, https://secure.historians.org/pubs/dissertations/. Data export provided to author by Liz Townsend, March 3, 2014.] Identifying the gender of their authors identifies the approximate number of women and men who are completing PhD-level training in history each year. Our program uses the year a historian wrote his or her dissertation to estimate a period for when they might have been born: we have assumed that a historian was between 25 and 45 years old when completing the dissertation. Using this temporal information, we are better able to infer their gender and chart how the larger representation of women and men changes year-by-year.

@@

Another way to examine this trend is to look at the proportion of male and female dissertation authors over time, which smoothes out changes in the absolute number of dissertations produced. The proportion of dissertations written by women has steadily increased over the past half-century, a change that began in the late 1960s and continued through the early 2000s. Since that point, the proportion of dissertations written by women has largely plateaued at a few percentage points below the proportion written by men. Female historians have achieved something approaching parity with male historians in terms of how many women and women complete dissertations each year.

@@

@@

[Goldstone and Underwood 2014] Goldstone, Andrew, and Ted Underwood. "The Quiet Transformations of Literary Studies: What Thirteen Thousand Scholars Could Tell Us." *New Literary History* (Forthcoming 2014).

@@

@@

@@

@@

>>>>

---
title: "Jane, John … Leslie? A Historical Method for Algorithmic Gender Prediction"
author: "Cameron Blevins and Lincoln Mullen"
output:
  html_document: default
  word_document: default
---

@@

This paper outlines a new method for inferring the gender of personal names in large, machine-readable datasets. In contrast to existing methods of predicting gender that treat names as static entities, this method uses a humanistic approach to take into account the dynamic nature of naming practices. It uses historical data to measure the likelihood that a name was associated with a particular gender based on the time or place under study. This approach generates more accurate results for sources that encompass changing periods of time, providing digital humanities scholars with a flexible tool to apply and adapt to analyses of gender. The article first describes the methodology as implemented in the [Gender](https://github.com/ropensci/gender) package for the [R programming language](http://www.r-project.org/). It goes on to apply the method to a case study in which we examine gender and gatekeeping in the American historical profession over the past half-century. The Gender package illustrates the importance of incorporating humanistic approaches to computer science and related fields.

@@

The analytical category of gender has transformed humanities research over the past half-century, and more recently, the digitization movement has opened up opportunities to apply this category to new sources and in new ways. Initiatives like the Orlando Project, the Poetess Archive, and the Women Writers Project have done crucial work to recover the voices of female authors and writers [Orlando Project, n.d., Poetess Archive n.d., Women Writers Project n.d.]. Other scholars have used digital analysis to study linguistic styles of male and female playwrights or representations of gendered bodies in European fairytales [Argamon et. al. 2009, Weingart and Jorgensen 2013]. The proliferation of machine-readable datasets and digital texts brings with it a vast trove of material with which to study gender. Much more, however, can still be done. To better equip humanities researchers to take full advantage of digital sources, we have developed a methodology that infers gender from one of the most common features of humanities datasets: personal names.

As humanities datasets grow larger and larger, researchers have not only been able to access information that is there, but to infer information that isn't there. Topic modeling, for example, has been used to uncover scholarship trends over a century of academic articles in literary studies. [Goldstone and Underwood 2014]. Can researchers similarly use personal names to infer information about individual people? After all, personal names appear across a wide range of sources in the humanities: literary characters in stories and novels, authorship information in archival metadata, or individuals mentioned by historical newspapers and magazine articles. Authors, characters, or historical actors frequently stand at the analytical center of humanities scholarship, and their names provide an intriguing way to infer additional information about them.^[The availability of large-scale datasets dealing with personal information raises important ethical issues for digital humanists, particularly in light of the revelations about domestic spying at the NSA brought forth by Edward Snowden in 2013. [Weingart 2014] Kieran Healy has used network analysis to demonstrate how researchers can infer information from datasets while also touching on the troubling ethical issue of surveillance [Healy 2013].]

One of the central kinds of information researchers can infer from a personal name is the gender of that individual: "Jane Fay," for instance, is almost certainly a woman, while her brother "John Fay" is almost certainly a man. A computer program can infer their respective genders by looking up “Jane” and “John” in an existing dataset that matches names to genders. But this approach runs into a problem: the link between gender and naming practices, like language itself, is not static. What about Jane and John Fay's sibling, Leslie? If Leslie was born in the past sixty years, chances are good that Leslie would be their sister. But if the three siblings were born in the early twentieth century, chances are good that Leslie would be their brother. That is because the conventional gender for the name Leslie switched over the course of one hundred years. In 1900 some 92% of the babies born in the United States who were named Leslie were male, while in 2000 about 96% of the Leslies born in that year were female.

@@

For those working on contemporary subjects the "Leslie problem" is not an especially pressing one. A film scholar analyzing male and female characters in modern American television shows, for instance, can turn to a variety of tools that use current databases of names, and these tools will likely identify the gender of Leslie Fay correctly. One online service, Genderize.io, determines the gender of first names "from user profiles across major social networks." [Genderize.io n.d.] Checking the service for the name Leslie returns a prediction that the name is female, along with an estimate of male/female proportions of the uses among members of unnamed social media sites.^[In this example the Genderize.io service has been accessed through an API wrapper provided in our Gender package described below.]

**I took out the genderize code here because I think it's not immediately intuitive for the reader, given that we haven't totally introduced our package yet. -CB**

But researchers studying a longer timespan need to take into account the changing nature of naming practices. If that same film scholar wanted to compare the characters of modern television shows to characters during the silent film era,  existing tools like Genderize.io would misidentify Leslie Fay's gender in the earlier screenplays. Other names used in the United States such as Madison, Morgan, Sydney, and Kendall, have also flipped their gender over the past century and would similarly be misidentified. The prevalence of changes in the gender of names in the last half of the twentieth century implies that gendered lists of names drawn from contemporary data are anachronistic and so of limited use for datasets that encompass changing periods of time.^[The few exceptions of names which flip from being female to male, such as "Jan", have a different causal mechanism. We speculate that Jan as an English name is female but that Jan as a Scandinavian name is male. Because the name has remained relatively uncommon, other causes besides changing gender preferences can explain the change.]

**I propose cutting this part or moving it to another section, as I think we need to move faster to get into the payoff of our method: Such changes usually follow set patterns. In general, names tend to shift from being predominantly male to being predominantly female; the shift rarely happens in the other direction. In general, names that transition from male to female also became increasingly popular over time. For example, the name Madison changed from male to female quite suddenly after 1985. Whereas before it had been an infrequently used male name, after 1985 it skyrocketed in popularity and by the early 2000s was one of the ten most frequently used names for baby girls. Broadly speaking, in the late twentieth century Americans girls became increasingly more likely than American boys to receive a "trendy" name, and once that name was associated with women it was unlikely to be given to male babies. These names present a moving target for researchers hoping to use them to infer additional information about the people under study.**

@@

The "Leslie problem" is not just for researchers who think of themselves as historians. Anyone studying a period longer than a few years, or anyone studying a group whose demographics do not match the groups used by a contemporary tool such as Genderize.io will also encounter this problem. For example, a literary scholar studying a corpus of poems from the mid-twentieth century might wish to compare the stanza structures of male and female poets. Using a tool like Genderize.io based on contemporary data risks misidentifying these writers, many of who were born (and named) more than a century prior to the creation of these modern name datasets. This problem will only increase with growing life expectancies: as of 2012, the average American lifespan was nearly seventy-nine years---more than enough time for naming practices to change quite dramatically between a person’s birth and death [World Bank 2014]. Predicting gender from first names therefore requires a dynamic method that takes into change over time.

Our solution to the "Leslie problem" is to create a software package that combines a predictive algorithm with several historical datasets suitable to various times and regions. The algorithm calculates the proportion of male or female uses of a name in a given birth year or range of years. It thus can provide not only a prediction of the gender of a name, but also a measure of that prediction's likelihood. For example, our program predicts that a person named Leslie born in 1950 is female, but only just barely: 52 percent of the babies named Leslie born in the United States in 1950 were girls and 48 percent were boys. Researchers can use these probabilities to determine what level of certainty is acceptable for predicting the gender of different names.

It is important to understand our method’s limitations as well as its advantages. The approach of inferring gender from personal names is still a blunt way to study a complex subject. Gender theorists and feminist scholars have spent decades unpacking the full range of meanings behind gender as an analytical concept. On its own, our method has little to say about crucial topics such as the intersectionality of gender with race, class, and power, or the fluid relationship between biological sex and individual identity [Butler 1990, Butler 1993, Crenshaw 1991, McCall 2005]. The tool itself does not reflect how an individual person identifies oneself and is neutral as to what the categories “male” or “female” actually mean in any given historical and social context. The question of gender identity is especially relevant for our method, given its reliance on large datasets created by state agencies such as the U.S. Census Bureau. Much like the state’s definition of racial categories, governments impose their own vision of a gender binary onto their subjects, which in turn exerts a tremendous definitional power over societies and individuals [Brubaker and Cooper 2000, 14-17, Pascoe 2009, Canaday 2011, Landrum 2014]. Researchers who use this method need to be fully aware of the ways it replicates this problematic, state-defined gender binary.

As with any tool, the onus is on the researcher to understand how its underlying data was gathered, what it implies, and how it reflects social and cultural practices. Imperfect as it may be, this method nevertheless gives digital humanities scholars an important means of studying gender in textual collections. We have made the method as transparent and flexible as possible through the inclusion of probability figures - ex. 52 percent of the babies named Leslie born in the United States in 1950 were girls and 48 percent were boys. This makes the underlying data visible to users and allows them the opportunity to interrogate the assumptions that come with it. When used thoughtfully, this method can infer additional information about individuals that stand at the heart of humanities scholarship, providing another way to see and interpret sources across large scales of analysis.

The remainder of this article is divided into two sections. First, we describe our method in more detail, outline its advantages over existing methods, and explain how to use it. Second, we apply the method to a case study of gatekeeping in the historical profession in order to demonstrate its usefulness for humanities scholars.

@@

In a 2005 report for the American Historical Association, Elizabeth Lunbeck acknowledged "a sea change in the [historical] profession with respect to gender" before going on to describe the painful limitations of this sea change for female historians: ongoing personal discrimination, lower salaries, and barriers to securing high-ranking positions. Lunbeck's report drew in part on a survey of 362 female historians that produced a rich source of responses detailing the deep and multi-faceted challenges facing women in the profession [Lunbeck 2005]. What follows is a quantitative supplement to Lunbeck's analysis that uses our program to analyze gender representation amongst historians across a much larger scale and a much longer time period. We study one of the bedrocks of the historical profession, scholarly research, and focus on two types of scholarly output: the dissertation and the monograph. Although women have achieved rough parity with men in terms of the number of completed dissertations, female historians continue to face an unequal playing field in the monographic landscape. Applying our method to the pages of the American Historical Review reveals a persistent and significant gap in the number of male-authored and female-authored books reviewed by one of the field’s premier academic journals.

We begin with the history dissertation, often the defining scholarly production of a historian's early career. The completion of a dissertation marks a standardized moment of transition out of the "training" phase of the historical profession. To identify how many women and men completed PhD-level training in history each year, we used data supplied by ProQuest for roughly eighty thousand PhD dissertations completed in the United States between 1950 and 2012.^[ProQuest's dataset does not permit us to limit dissertations just to dissertations produced in history departments. These dissertations include dissertations *about* history. For a justification of this approach, see a [series of blog posts](http://lincolnmullen.com/research/history-dissertations/) exploring the dataset. The American Historical Association's dataset of dissertations relies on reporting from departments and does not have the chronological range of the ProQuest data. Nevertheless, applying the same method to the AHA's dataset does not provide any reason for thinking that the proportion of authorship by gender over time is substantially different when limited to just history department dissertations. American Historical Association, *Directory of History Dissertations*, https://secure.historians.org/pubs/dissertations/. Data export provided to author by Liz Townsend, March 3, 2014.] Our program used the year a historian wrote his or her dissertation to estimate a period for when they might have been born, assuming that a historian was between 25 and 45 years old when completing their dissertation. With this temporal information, we were able to infer their gender and chart how the larger representation of women and men changed year-by-year.

@@

Another way to examine this trend is to ignore changes in the *absolute* number of dissertations produced each year and to instead look at changes in the *proportion* of male and female dissertation authors. The proportion of dissertations written by women has steadily increased over the past half-century, a change that began in the late 1960s and continued through the early 2000s. Since that point, the proportion of dissertations written by women has largely plateaued at a few percentage points below the proportion written by men. Female historians have achieved something approaching parity with male historians in terms of how many women and men complete dissertations each year.

@@

[Crenshaw 1991] Crenshaw, Kimberle. "Mapping the Margins: Intersectionality, Identity Politics, and Violence against Women of Color." *Stanford Law Review*. 43.6 (July 1991): 1241-1299.

@@

[Goldstone and Underwood 2014] Goldstone, Andrew, and Ted Underwood. "The Quiet Transformations of Literary Studies: What Thirteen Thousand Scholars Could Tell Us." *New Literary History* 45.3 (Summer 2014): 359-384.

@@

[McCall 2005] McCall, Leslie. "The Complexity of Intersectionality." *Signs*. 30.3 (Spring 2005): 1771-1800.

@@

[Orlando Project n.d.] Orlando Project. "About." <http://www.artsrn.ualberta.ca/orlando/?page_id=2>. Accessed 23 June 2015.]

@@

[The Poetess Archive n.d.] The Poetess Archive. "About the Poetess Archive." <http://idhmc.tamu.edu/poetess/about/index.html>. Accessed 23 June 2015.]

@@

[Weingart and Jorgensen 2013] Weingart, Scott, and Jeana Jorgensen. "Computational Analysis of the Body in European Fairy Tales." *Literary and Linguistic Computing*, 28.3 (2013): 404-416.
